**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-03-16 Manuscripts/Manuscript_AM_v4.Rmd
Working on v4 of manuscript
Updated final two sections, need to work on figure of linear and quadratic response

A linear mixed model for a longitudinal study can be written as:

@@

When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is apparent that the LMEM and rm-ANOVA models have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$). This term is defined as a _random effect_, which accounts for variability in each subject within each group. The _random_ component can be defined as "noise" caused by some inherent variability within the study that is not captured by the term $\varepsilon_{tij}$.

@@

@@

>>>>

A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014]. A LMEM with interaction between time and treatment for a longitudinal study  can be written as:

@@

When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$). This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group. The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{tij}$ from Equation \@ref(eq:linear-model).

@@

Although the specific methods of how GAMs model covariance and correlation is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in [@hefley2017]. In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time. An example where this is covered is in the Appendix.

@@

At the core of a biomedical longitudinal study lies the question of a significant difference between two or more treatments in different groups. Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_, GAMs use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs by group is that if the _difference_ between the confidence intervals of the smooths is not equal to zero then a significant difference exists at that timepoint (or timepoints). The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent. The estimation of siginificant differences in GAMs is covered in detail in the Appendix.
